---
title: "Getting Started"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>"
)
```

`MiDAR` is an R package to reproducibly manage, post-process, visualize, apply quality control, and analyze small-molecule mass spectrometry (MS) datasets, e.g. from targeted lipidomics and metabolomics experiments.

`MiDAR` is tailored to handle different analytical designs, data types and data processing strategies. As such, this package provides functions to import data files from from different commercial and open-source tools. Data processing functions include, internal standard and sample amount-based normalization, quantification, as well as drift and batch corrections. Quality control (QC) functions provide QC metrics and plots of raw and processed data, and QC-based feature filtering.  

Datasets and processing steps are tracked, and can be saved as `MiDAR` S4 class object (RDS) files, Excel, PowerPoint and interactive HTML-based reports, enabling sharing and reporting of all data, metadata and applied data processing steps.

## Example Workflow

Below is example of a simple data processing workflow, starting with a MassHunter CSV file, metadata in an Excel template 

```{r}
library(midar)

# Get paths of example files included with this package
masshunter_file <- system.file("extdata", "Example_MHQuant_1.csv", package = "midar", mustWork = TRUE)
metadata_file <- system.file("extdata", "Example_Metadata_1.xlsm", package = "midar", mustWork = TRUE)

0# Create a MidarExperiment object (S4)
mexp <- MidarExperiment()

# Load data and metadata
mexp <- midar::import_data_masshunter(data = mexp, file_dir_names = masshunter_file, file_format = "csv")
mexp <- midar::import_metadata_msorganizer(data = mexp, filename = metadata_file, excl_unannotated_analyses = TRUE)

# Normalize and quantitate each feature by internal standards
mexp <- normalize_by_istd(mexp)
mexp <- quantitate_by_istd(mexp)

# Get QC metrics for each feature
mexp <- calculate_qc_metrics(mexp)

# Filter features according to QC criteria
mexp <- apply_qc_filter(data = mexp,
                        exclude_technical_outliers = TRUE,
                        min_cv_conc_bqc = 30,
                        min_intensity_bqc = 100,
                        min_signal_blank_ratio = 5,
                        min_response_rsquare = 0.8,
                        rqc_curve_used_for_filt = 1)



```

## Contributor Code of Conduct

Please note that the midar project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
