---
title: "Getting Started with MiDAR"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  MiDAR: Postprocessing and Quality Control of Small-Molecule Mass Spectrometry Data 
vignette: >
  %\VignetteIndexEntry{Getting Started with MiDAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```
# Prerequisites
`MiDAR` requires [R](https://cran.r-project.org) version 4.2.0 or higher) to run. 
Usage of an R IDE (integrated development environment) such as [RStudio](https://posit.co/download/rstudio-desktop/), [Positron](https://positron.posit.co), 
or [Visual Studio Code](https://code.visualstudio.com) is also recommended.

While `MiDAR`'s core functionalityr requires no R coding skills,
basic familiarity with R and an IDE is needed. New to R? Try online tutorials such a
[An opinionated tour of RStudio](https://rladiessydney.org/courses/ryouwithme/01-basicbasics-1/) and 
the [RStudio User Guide](https://docs.posit.co/ide/user/ide/get-started/). Additionally,
having a colleague who is familiar with RStudio can be helpful
to get started smoothly and stay motivated.

# Installing MiDAR
To install, or to update, `MiDAR`, run the following code in the R console:
```{r, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pkg_install("SLINGhub/midar")
```

# Learning MiDAR
If you are new to MiDAR, please explore the tutorials available via the
top menu bar. These resources will help you become familiar with the
package's concepts and functionalities. In particular, the tutorials on 
[Preparing and importing data](T01_prepdata.html) and [A basic MiDAR workflow](T02_settingup_workflow.html) offer essential information 
to get you started smoothly.

Some MiDAR functions, particularly those for plotting, have many arguments
to allow detailed customizations. To get started,  use the default
settings (i.e., without defining them), and then
adjust them to suit your needs. Use `?functionname` or search the [References(reference/index.html)
page for detailed descriptions, or check the 'Manual' section for tips and
tricks. A common error is omitting required arguments. If unclear errors occur,
consult the Help to ensure correct arguments and data types. Robustness
will improve in future package versions.

# MiDAR Recipes
To build your own data processing workflows, the 'Recipes' provide a good
starting point. Choose one or more recipes that match your application, copy-paste into your
R script/notebook, and start from there to adapt to your specific data and process. 
The recipes also introduce you to some function names and argument
options that may typically be used in the recipes' context. 

For more detailed information, refer to the 'Manual' section for
details on data structures and the use of MiDAR functions, and the
'Reference' section for comprehensive documentation on all functions,
their arguments, data classes, and test datasets.

# Support
We are happy to help. Feel free to contact the authors directly or via the Github repository for any questions, suggestions, or issues. If you are interested in contributing to the package, feel free to contact us.


