---
title: "Importing analytical data"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  MiDAR: postprocessing and quality control of small molecule mass spectrometry data 
vignette: >
  %\VignetteIndexEntry{Importing analytical data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Analytical data, i.e. preprocessed data from mass spectrometry
experiments, can be imported from different sources.

Data files present in a folder can also be imported and merged as well.
This can be useful when the raw data processing is broken down in
batches resulting in separate result files.

# Data Sources

Following formats are currently supported:

| Source | MiDAR function | Details | File |
|--------------|----------------|-----------------------------|--------------|
| Agilent MassHunter | `import_data_massshunter()` | Flat and nested tables from MassHunter Quant. | .csv |
| MRMkit | `import_data_mrmkit()` | Long format output format | .tsv |
| Plain wide CSV | `import_data_table()` | Samples/analyses as rows and features as columns. Can contain columns specific sample annotations. | .csv |

# Metadata within analytical results

When the analytical results contain metadata, such as sample and feature
annotations, these can be imported as metadata in the `MidarExperiment`
object as well. The imported metadata is checked for integrity and
consistency (see TODO) and then added to the annotation tables within
the `MidarExperiment`. To include available metadata, set the argument
`import_metadata = TRUE`.

# MRMkit Results

Output files from
[MRMkit](https://github.com/MRMkit/MRMkit "https://github.com/MRMkit/MRMkit"),
an open-source peak integration software for MRM data
(\cite{author2023}) can imported directly. Specific metadata present in
the data file can be imported as well (\`import_metadata = TRUE\`)

```{r}
library(midar)
filepath <- system.file("extdata/MRMkit_demo.tsv", package = "midar")
myexp <- MidarExperiment()

myexp <- import_data_mrmkit(myexp, filepath, import_metadata = TRUE)
```

# Agilent MassHunter Quantitative

Peak integration results exported from Agilent Masshunter Quant in the
CSV format can be imported. Samples must be present in rows, features as
columns. Import of qualifier results is supported. Sample, method and
result metadata present in the files can also be imported
(\`import_metadata = TRUE\`)

```{r}
filepath <- system.file("extdata/MHQuant_demo.csv", package = "midar")
myexp <- MidarExperiment()

myexp <- import_data_masshunter(myexp, filepath, import_metadata = TRUE)
```

# Plain CSV files

Analysis results, whether raw intensities (e.g., peak areas) or
preprocessed data (e.g., concentrations), can be provided as plain CSV
tables. In these tables, analyses (samples) should be arranged in rows,
and features in columns. The specific data type in the table (e.g., area
or concentration) is defined using the `variable_name` argument.

```{r}
filepath <- system.file("extdata/plain_wide_dataset.csv", package = "midar")
myexp <- MidarExperiment()

myexp <- midar::import_data_csv(
  myexp, path = filepath,
  variable_name = "area", 
  import_metadata = TRUE)
```

# Importing and merging multiple files

Multiple data files can be imported and merged. Users can
either provide a list of file paths or specify a folder path to import
all data files within that directory. This support for multiple files is
useful when raw data processing is divided into batches, leading to
separate result files.

The imported merged data is checked for consistency, ensuring that each
analysis ID and feature ID pair is unique. This means that the same
feature cannot be reported multiple times within the same analysis,
which can happend for example if the same feature in the same sample was
integrated in different raw data processing batches.
