---
title: "Metadata import"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  MiDAR: postprocessing and quality control of small molecule mass spectrometry data 
vignette: >
  %\VignetteIndexEntry{Metadata import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)

if(fs::file_exists("metadata_template.xlsx")) fs::file_delete("metadata_template.xlsx")
if(fs::file_exists("metadata_msorganiser_template.xlsm")) fs::file_delete("metadata_msorganiser_template.xlsm")
```

### Metadata

Metadata in this context refers to analysis metadata, i.e., data that
annotate the analytical data. Metadata can be retrieved from the
imported analysis data file as far as available, or can be imported from
separate files or R data frames.

Integrity of metadata and data is instrumental for correct and smooth
post processing of data. Midar therefore inspects imported data/metadata
for completness of data and consistency of IDs used across differet
metadata tables. After importing metadata a summary of identified error,
warnings and notes cocerning metadata is printed to the console, to
allow user to identify and adress possible issues.

### Metadata formats and temples

The structure and required/optional columns for each metadata type is
descrived in the Manual and in the help of the correspoding import
functions.

To obtain templates for metadata, a Excel file containing all template
metadata table templates can be saved.

```{r}
midar::save_metadata_templates()
```

### Importing metadata from files/sheets

First we import the analysis data as outlined in "Importing data". In
this case we explicitly do not not import any metadata present in the
analysis data, i.e., we only import the peak areas.

```{r}
library(midar)
mexp <- midar::MidarExperiment()

data_path <- "datasets/sPerfect_MRMkit.tsv"
mexp <- import_data_mrmkit(data = mexp, path = data_path, import_metadata = TRUE)
```

Now we can add corresponding metadata file-by-file

```{r}
mexp <- import_metadata_analyses(mexp, 
                                 path = "datasets/analysis_metadata.csv", 
                                 excl_unmatched_analyses = TRUE, 
                                 ignore_warnings = TRUE)

mexp <- import_metadata_features(mexp, 
                                 path = "datasets/feature_metadata.csv",
                                 ignore_warnings= TRUE )
```

Metadata can also be imported from sheets of an Excel workbook, which
allows to store all metadata in one file. \
In this case below we add metadata on internal standard and response
curves to the MidarExperiment object

```{r}
mexp <- import_metadata_istds(mexp, 
                              path = "datasets/metadata_tables.xlsx", 
                              sheet = "ISTDs",
                              ignore_warnings= TRUE)
```

Furthermore, metadata can be imported from R `data.frame` objects, thus
allowing users to obtain metadata from additional sources, e.g. from
databases or a LIMS.

```{r}
df_qcinfo <- readr::read_table(file = "datasets/qc_metadata.txt")
mexp <- import_metadata_qcconcentrations(mexp, table = df_qcinfo)
```

### Importing MSOrganiser metedata file template

Another option to import metadata is via the MSOrganiser template file,
a macro-based (XLSM) Excel file. This template offers tables for all
metadata types supported by MiDAR, with options to do perform checks on
validatiy and integrity of the metadata.

The template can be obtained from <https://github.com/SLINGhub/midar> or
via a midar function

```{r}
midar::save_metadata_msorganiser_template()
```

Only the metadata tables required by the intended processing workflow
need to be completed. With following import function all completed
tables will be imported.

```{r}
mexp <- import_metadata_msorganiser(mexp, 
                                    path = "datasets/sPerfect_Metadata.xlsm",
                                    ignore_warnings= TRUE)
```

### 
