---
title: "A basic MiDAR workflow"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  MiDAR: Postprocessing and Quality Control of Small-Molecule Mass Spectrometry Data 
vignette: >
  %\VignetteIndexEntry{A basic MiDAR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = TRUE, message = TRUE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

This tutorial outlines some key steps in a MiDAR workflow, based on a lipidomics dataset.
See the tutorial [Lipidomics Data Processing](articles/T01_targetlipidomics_workflow.html) for a more detailed example.

Keep in mind that these examples below are simplified and may not be applicable to your data and experimental setup.
Please consult other tutorials and recipes for information on other workflows and data types.

# Setting up a RStudio project

To start a new MiDAR data analysis, creating a RStudio project is recommended. 
(See [Using RStudio Projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)).
This will help to keep the data analysis organized and makes it easier to share with others. 
The project should contain the following subfolders: `data` and `output`. Add data and metadata files to the `data` folder.

Notebooks such as R/Notebook (.rmd) or [Quarto Notebook](https://docs.posit.co/ide/user/ide/guide/documents/quarto-project.html) (.qmd) are 
good choices to create documented data processing workflows. These formats allow 
combining code with formatted text to document the data processing steps.

Start with a new notebook or R script and load the `midar` package

```{r}
library(midar)
```

# Creating a MidarExperiment object
The `MidarExperiment` object is the main **data container** used in the MiDAR workflow. 
See (The `MidarExperiment` data object)[articles/03_midarexperiment.html] for more information.

We start by creating a new `MidarExperiment` object (`myexp`), which will be used 
in all subsequent steps.

```{r}
myexp <- MidarExperiment()
```


# Importing analysis results

As introduced in [Preparing and importing data](articles/T01_prepdata.html)  
we first import the analytical data, in this case from [MRMkit](https://github.com/MRMkit/MRMkit) file.
This file also contains some metadata, such as `qc_type` and `batch_id` (see [Data Identifiers in MiDAR](articles/02_keydataids.html)) 
which we will import as well.

```{r load-rawdata}
myexp <- import_data_mrmkit(data = myexp, 
                            path = "datasets/sPerfect_MRMkit.tsv", 
                            import_metadata = TRUE)
```

# Adding metadata

The subsequent processing steps require additional infomation that is not available
from the imported analysis data. This includes information such as which internal standards are used to normalize, their concentrations and the sample amounts analysed.
These metadata can be imported from separate files or R data frames as 
described in [Preparing and importing data](articles/T01_prepdata.html).
To keep the code concience in this example, we will import metadata from an msorganiser
template. The validation checks result in some warnings, which will by default result in a failed metadata import.
However, assuming we understand what we are doing, 
we decided to ignore these warnings by setting `ignore_warnings = TRUE`. They will still be shown
in the table printed in the console, labelled with an asterix (*) in the `status` column.

```{r import-metadata}
myexp <- import_metadata_msorganiser(
  myexp, 
  path = "datasets/sPerfect_Metadata.xlsm", 
  ignore_warnings = TRUE
)

```

# Applying Data Processing
Now we are ready to proceed with data processing. In this example we will 
employ some basic data processing steps, whereby the corresponding code should be 
self explanatory. At the end of this code block we also set criteria based on which 
features are filtered on demand later in the workflow.

```{r processinh}
myexp <- normalize_by_istd(myexp)
myexp <- quantify_by_istd(myexp)

myexp <- correct_drift_gaussiankernel(
  data = myexp,
  variable = "conc",
  ref_qc_types = c("SPL"))

myexp <- midar::correct_batch_centering(
  myexp, 
  variable = "conc",
  ref_qc_types = "SPL")

myexp <- filter_features_qc(
  data = myexp,
  include_qualifier = FALSE,
  include_istd = FALSE, 
  min.signalblank.median.spl.sblk = 10,
  max.cv.conc.bqc = 25)
```

# Plotting data
MiDAR provides various plot function that can be useful in understanding analytical 
performance, trends and isses in the data. Plots are also available to inspect the 
effect of data processing (e.g. drift/bath-effect correction) and QC-based feature filtering.

Below we create a `runscatter` plot to visualize the concentration of specific
features across the analytical series in different QC sample types. 
The plot can also be saved to a PDF file. 

```{r}
#| fig.alt: >
#|   RunScatter plot with 2x3 panels
  plot_runscatter(
    data = myexp,
    variable = "conc",
    include_feature_filter = "PC 4",
    include_istd = FALSE,
    cap_outliers = TRUE,
    log_scale = FALSE,
    output_pdf = FALSE,
    path = "./output/runscatter_PC408_beforecorr.pdf",
    cols_page = 3, rows_page = 2,
  )
```


# Exporting and sharing data
Finally, we can export specific datasets as plain csv tables, create a detailed data report, 
and share the entire `MidarExperiment` object with someone else without any code, 
who can run own data processing, plots and QC checks.

```{r save-report}
# Saves a detailed report in Excel format with multiple sheets
midar::save_report_xlsx(myexp, path = "./output/myexp-midar_report.xlsx")

# Saves flat csv table with concentration values that passed the previously set
# QC criteria, for each feature in each sample. 
midar::save_dataset_csv(
  data = myexp, 
  path = "./output/sperfect_filt_uM.csv",
  variable = "conc", 
  qc_types = "SPL", 
  include_qualifier = FALSE,
  filter_data = TRUE)

# Saves the entire MidarExperiment object as an RDS file, which can be
# opened in R without MiDAR or used with MiDAR again.
saveRDS(myexp, file = "./output/myexp-midar.rds", compress = TRUE)
```

