---
title: "Batch-effect correction"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Batch-effect correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

MiDAR currently supports median centering-based batch effect correction.

The data must be provided via a MidarExperiment object, whereby raw data that was imported or
processed data can be corrected, such as `intensity` or `conc` values.

## Import data

In this tutorial, we import pre-calculcated concentration values from a CSV file. This file must
contain a column with batch information (`batch_id`), see `import_data_csv()` for more information.

```{r setup}
#| results: false  
#| message: false
library(midar)

myexp <- midar::MidarExperiment()

myexp <- import_data_csv(
  myexp,
  path = "simdata-u1000-sd100_7batches.csv", 
  variable_name = "conc", 
  import_metadata = TRUE)
```

## Apply batch-centering

Now we center the concentration values of batches based on a reference qc type, the samples
(`ref_qc_types = "SPL"`) in this case. The `correct_scale` parameter

```{r}
#| results: false  
#| code-overflow: 'wrap'

myexp <- correct_batch_centering(
  data = myexp, 
  variable = "conc",
  ref_qc_types = "SPL", 
  correct_scale = FALSE
)

```

Next, we inspect the data before and after batch correction. We observe that the batches are now
aligned. Note: If the samples or other quality control types do not follow the reference samples,
they may not be appropriately corrected.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
plot_runscatter(myexp, variable = "conc_before", rows_page = 1, cols_page = 1)
plot_runscatter(myexp, variable = "conc", rows_page = 1, cols_page = 1)
```

## Batch-centering with variance scaling

Above, we observe that while the batches are aligned, the spread (variance) of the data points
varies considerably between the batches. We can correct this by scaling the variance via
`correct_scale = TRUE`. As shown in the plot below, the variance of the data points now appears
fairly consistent across the batches.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
myexp <- midar::correct_batch_centering(
  myexp, 
  ref_qc_types = "SPL", 
  variable = "conc",
  correct_scale = TRUE
)

plot_runscatter(myexp, variable = "conc", rows_page = 1, cols_page = 1)

```

## Export batch-corrected data

Next, we can either continue to work with the corrected data using `MiDAR` functions or export the
data.

```{r}
save_dataset_csv(
  myexp, 
  path = "batch-corrected-conc-data.csv", 
  variable = "conc", 
  filter_data = FALSE
  )
```
