---
title: "Calibration by a Reference Sample"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Calibration by a Reference Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

library(tidyverse)
library(gt)
```

Feature abundances in samples can also be calibrated to corresponding
abundances in a specified reference sample. MiDAR supports absolute
(re-)calibration and normalization (relative calibration).

Absolute calibration of feature abundances is based on known metabolite
concentrations in a reference sample (e.g., NIST SRM1950 plasma).
Normalization (relative calibration) is based on calculating the abundance ratios of features in
samples and a reference sample.

Below, we will demonstrate both absolute and relative calibration using NIST
SRM1950 plasma samples that were measured as part of the same analysis..

## Import data and metata,

```{r setup}
#| results: false  
#| message: false
library(midar)

# Get example data paths
dat_file = system.file("extdata", "S1P_MHQuant.csv", package = "midar")
meta_file = system.file("extdata", "S1P_metadata_tables.xlsx", package = "midar")

# Load data and metadata
mexp <- MidarExperiment()
mexp <- import_data_masshunter(mexp, dat_file, import_metadata = FALSE)
mexp <- import_metadata_analyses(mexp, path = meta_file, sheet = "Analyses")
mexp <- import_metadata_features(mexp, path = meta_file, sheet = "Features")
mexp <- import_metadata_istds(mexp, path = meta_file, sheet = "ISTDs")
```

## Load known analyte concentrations of the reference sample

A table containing the known analyte concentrations for the NIST SRM1950 reference 
sample is now been added to the `MidarExperiment object`. 
Please note that the S1P concentrations provided in this table are intended for 
illustrative purposes only. The actual absolute S1P concentrations in NIST SRM1950 
may differ significantly.

```{r}
#| results: false  
#| code-overflow: 'wrap'
#| message: true

mexp <- import_metadata_qcconcentrations(mexp, path = meta_file, sheet = "QCconcentrations")

```

## Process the data

The analysis was performed using HILIC chromatography; thus, we need to
correct for the isotope interferences from S1P 18:2;O2 M+2 and S1P 18:1;O2 M+2.
Subsequently, initial quantification is done using the spiked-in ISTD concentration.

```{r}
#| results: false 
#| message: true

# Isotope correction
mexp <- midar::correct_interferences(mexp)

# Quantify the data
mexp <- normalize_by_istd(mexp)
mexp <- quantify_by_istd(mexp)
```

## Absolute calibration

We perform the absolute re-calibration using the function
`calibrate_by_reference()`. The reference sample is set via
`reference_sample_id`. In cases where multiple analyses of the same reference
sample are present in the dataset, either the mean or median is calculated
(defined via the `summarize_fun`).

The calibrated concentration is calculated as:

$$
c_\text{calibrated}^\text{Analyte} = \frac{c_\text{sample}^\text{Analyte}}{c_\text{ref}^\text{Analyte}} \times c_\text{known}^\text{Analyte}
$$
```{r}
#| results: false 
#| message: true
 
mexp_res <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = TRUE,
    batch_wise = FALSE,
    summarize_fun = "mean",
    undefined_conc_action = "na"
  )
```

The re-calibrated concentrations are stored in the variable `conc` , overwriting
any previously calculated concentrations. The original concentrations, however,
are still available via the variable `conc_beforecal`. 

The re-calibrated concentrations can be exported as usual, and they also
appear in the MiDAR XLSX report as concentrations.

```{r}
#| results: false 
#| message: true
# Export absolute calibration concentrations
save_dataset_csv(mexp, "calibrated.csv", variable = "conc")
  
# Export non-calibrated concentrations
save_dataset_csv(mexp_res, "noncalibrated.csv", variable = "conc_beforecal")

# Create XLSX report with calibrated concentrations as filtered dataset
save_report_xlsx(mexp_res, "report.xlsx", filtered_variable = "conc")
```

## Normalization (relative calibration)

We can perform a simple normalization with a reference sample using the
`calibrate_by_reference()` function, setting `absolute_calibration = FALSE`.
In this approach, in cases where multiple analyses of the same reference
sample are present in the dataset, either the mean or median is calculated
(defined via the `summarize_fun`).

```{r}
#| results: false 
#| message: true
 
mexp_res <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = FALSE,
    summarize_fun = "mean"
  )
```

The results of the normalization are stored, unlike for the absolute
calibration, as **ratios**, in a new variable, `[VARIABLE]_normalized`, where
[VARIABLE] is the input variable, e.g., `conc_normalized` or
`intensity_normalized`.

The normalized concentrations can be exported as [VARIABLE]_normalized using
`save_dataset_csv()`. In the MiDAR XLSX report generated by `save_report_xlsx()`,
the unfiltered dataset with normalized concentrations is included by default.
To include the normalized concentrations as the filtered dataset, set
`filtered_variable = “[VARIABLE]_normalized` as an argument.

```{r}
#| results: false 
#| message: true
# Export NIST1950-normalized concentrations
save_dataset_csv(mexp_res, "norm.csv", variable = "conc_normalized")

# Create XLSX report with normalized concentrations as filtered dataset
save_report_xlsx(mexp_res, "report_norm.xlsx", filtered_variable = "conc_normalized")
```

## Batch-wise calibration

Calibration can also be applied batch-wise, in which case each batch is
calibrated separately with the data from the reference sample in the same
batch. This is done by setting `batch_wise = TRUE` and can be used in both
absolute and relative calibration.

This approach can be used to correct batches or assays/plates using a reference
material shared across the batches.

```{r}
#| results: false 
#| message: true
 
mexp_res <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = TRUE,
    batch_wise = TRUE,
    summarize_fun = "mean",
    undefined_conc_action = "na"
  )

save_dataset_csv(mexp_res, "bathwise_calibrated.csv", variable = "conc_beforecal")
```

## Concentration ratio and bias

To examine the ratio between measured and expected (known) concentrations in the reference samples, a table with concentration ratios can be generated using the code below.

The ratio is calculated as follows:

$$
R_\text{ratio}^\text{Analyte} = \frac{c_\text{measured}^\text{Analyte}}{c_\text{expected}^\text{Analyte}}
$$

```{r}
#| results: true 
#| message: true

mexp_res <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = TRUE,
    summarize_fun = "mean",
    undefined_conc_action = "na",
    store_conc_ratio = TRUE
  )

tbl_ref_bias <- mexp_res$dataset |> 
  filter(sample_id == "SRM1950", is_quantifier) |> 
  group_by(feature_id) |> 
  summarise(bias_mean = mean(feature_conc_ratio))

gt::gt(tbl_ref_bias)


```
A ratio value of 1 indicates perfect agreement between the measured and expected concentrations.  
Values greater than 1 suggest overestimation, while values less than 1 indicate underestimation.  
The ratio values can also be visualized or further analyzed to identify outliers or investigate potential issues in the analytical process or calibration.

We can also calculate ratio and bias of QC samples directly without haveing to apply `calibrate_by_reference()`. 
For illustration, we calculated the ratio and bias of the QC samples using the re-calibrated data from above, expecting
the all corrected feature concentrations in the reference sample to have no bias (0%) and a ratio of 1.


```{r}
#| results: true 
#| message: true

tbl <- get_qc_bias_variability(mexp_res, qc_types = "NIST")
gt::gt(tbl) |> gt::fmt_number(decimals = 3)

```

The bias and concentration rations before the re-calibration can be viewed by using 
the `MidarExperient` object that had no calibration applied

```{r}
#| results: true 
#| message: true

tbl <- get_qc_bias_variability(mexp, qc_types = "NIST", with_conc_ratio = TRUE)
gt::gt(tbl) |> gt::fmt_number(decimals = 3)

```
