---
title: "Calibration by a Reference Sample"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Calibration by a Reference Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

Feature abundances in samples can also be calibrated to corresponding abundances in specified
reference sample. `MiDAR` supports absolute (re-)calibration and normalization (relative
calibration).

Absolute calibration of feature abundances is based on known metabolite concentrations in a
reference sample (e.g. NIST SRM1950 plasma). Normalization is based on calculateing the abundance
rations of feature in samples and in a reference sample.

Below, we will demonstrate both absolute and relative calibration using NIST SRM 1950 plasma samples
that were measured as part of the same analyses.

## Import data and metata,

```{r setup}
#| results: false  
#| message: false
library(midar)

# Get example data paths
dat_file = system.file("extdata", "S1P_MHQuant.csv", package = "midar")
meta_file = system.file("extdata", "S1P_metadata_tables.xlsx", package = "midar")

# Load data and metadata
mexp <- MidarExperiment()
mexp <- import_data_masshunter(mexp, dat_file, import_metadata = FALSE)
mexp <- import_metadata_analyses(mexp, path = meta_file, sheet = "Analyses")
mexp <- import_metadata_features(mexp, path = meta_file, sheet = "Features")
mexp <- import_metadata_istds(mexp, path = meta_file, sheet = "ISTDs")
```

## Load know analyte concentrations of the reference sample (NIST SRM1950)

Note that these S1P concentrations are based on an a different analysis and only meant as example
data. Actual absolute S1P concentrations of the NIST SRM1950 may considerably vary.

```{r}
#| results: false  
#| code-overflow: 'wrap'
#| message: true

mexp <- import_metadata_qcconcentrations(mexp, path = meta_file, sheet = "QCconcentrations")

```

## Process the data

The analysis was performed using HILIC chromatography, thus we need to correct for the isotope
interferences from S1P 18:2;O2 M+2 and S1P 18:1;O2 M+2. Initial quantification is done using
spiked-in ISTD concentration.

```{r}
#| results: false 
#| message: true

# Isotope correction
mexp <- midar::correct_interferences(mexp)

# Quantify the data
mexp <- normalize_by_istd(mexp)
mexp <- quantify_by_istd(mexp)
```

## Absolute calibration

We perform the absolute re-calibration using the function `calibrate_by_reference()`. The reference
sample is set via `reference_sample_id`. In case multiple analysis of the same reference sample are
present in the dataset, either the `mean` or `median` is calculated (defined via the
`summarize_fun`)

```{r}
#| results: false 
#| message: true
 
mexp <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = TRUE,
    batch_wise = FALSE,
    summarize_fun = "mean",
    undefined_conc_action = "na"
  )
```

The re-calibrated concentrations are stored in the `conc` variable, overwriting any previously
calculated concentrations. The original concentrations, however, are still available via
`conc_beforecal` variable.

The re-calibrated concentrations can be exported as usual and they also appear in MiDAR XLSX report
also as concentrations.

```{r}
#| results: false 
#| message: true
# Export absolute calibration concentrations
save_dataset_csv(mexp, "calibrated.csv", variable = "conc")
  
# Export non-calibrated concentrations
save_dataset_csv(mexp, "noncalibrated.csv", variable = "conc_beforecal")

# Create XLSX report with calibrated concentrations as filtered dataset
save_report_xlsx(mexp, "report.xlsx", filtered_variable = "conc")
```

## Normalization (relative calibration)

We can perform a simple normalization with a reference sample also using the
`calibrate_by_reference()` function, with setting `absolute_calibration = FALSE`. Also in this
approach, in case of multiple analysis of the same reference sample present in the dataset, either
the `mean` or `median` is calculated (defined via the `summarize_fun`)

```{r}
#| results: false 
#| message: true
 
mexp <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = FALSE,
    summarize_fun = "mean"
  )
```

The results of the normalization is stored, unlike for the absolute calibration, as **ratios**, in a
new variable, `[VARIABLE]_normalized`, whereby `[VARIABLE]` is the input variable, e.g.,
`conc_normalized` or `intensity_normalized`.

The normalized concentrations can be exported as `[VARIABLE]_normalized` using `save_dataset_csv()`.
In the MiDAR XLSX report generated by `save_report_xlsx()` the unfiltered dataset with normalized concentrations 
is included by default. If include the normalized concentrations as the filtered dataset, set
`filtered_variable = "[VARIABLE]_normalized" as argument.

```{r}
#| results: false 
#| message: true
# Export NIST1950-normalized concentrations
save_dataset_csv(mexp, "norm.csv", variable = "conc_normalized")

# Create XLSX report with normalized concentrations as filtered dataset
save_report_xlsx(mexp, "report_norm.xlsx", filtered_variable = "conc_normalized")
```

## Batch-wise calibration

Calibration can also be applied batch-wise, in which case each batch is calibrated 
separately with the data from the reference sample in the same batch. This is done 
by setting `batch_wise = TRUE` and can be used in both absolute and relative calibration.

This approach can be used to correct batches or assays/plates using a reference material 
shared across the batches.


```{r}
#| results: false 
#| message: true
 
mexp <- calibrate_by_reference(
    data = mexp,
    variable = "conc",
    reference_sample_id = "SRM1950",
    absolute_calibration = TRUE,
    batch_wise = TRUE,
    summarize_fun = "mean",
    undefined_conc_action = "na"
  )

save_dataset_csv(mexp, "bath-wise calibrated.csv", variable = "conc_beforecal")
```


