---
title: "Run-order Drift Correction (Smoothing)"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Run-order Drift Correction (Smoothing)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

MiDAR provides different functions for run-order drift correction, differing by the employed
smoothing algorithm. These functions can be categorized into functions that are either more suited
for QC or study sample-based drift correction.

The data must be provided via a MidarExperiment object, whereby raw data that was imported or
processed data can be corrected, such as `intensity` or `conc` values.

These drift correction functions have various options for customization, please refer the manual
page on [Drift and Batch Correction](07_driftbatchcorr.html) for more details.

## Import data

In this tutorial, we import pre-calculcated raw oncentration values from a CSV file. This file must
contain a column with batch information (`batch_id`) if batch-wise correction should be applied, see
`import_data_csv()` for more information.

```{r setup}
#| results: false  
#| message: false
library(midar)

myexp <- midar::MidarExperiment()

myexp <- import_data_csv(
  myexp,
  path = "smooth-testdata.csv", 
  variable_name = "conc", 
  import_metadata = TRUE)
```

## QC-based smoothing

Now we apply a QC-based drift correction using cubic spline. See `correct_drift_cubicspline()` for
more information.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_cubicspline(
  myexp,
  batch_wise = FALSE,
  variable = "conc",
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE)
```

Next, we inspect the data before and after the correction. We observe that the green trendline
follows the QC samples (in red). After the correction, the trend has been fully smoothed out,
resulting in a straight line.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

## Sample trend-based smoothing

Above, we observe that the QC samples do not fully represent the trends of the study samples. This
discrepancy can occur when the QC samples, which often based on pooled samples, differ in handling
or properties from the study samples.

We therefore now try a sample-based drift correction gaussian kernel smoothing. See
`correct_drift_gaussiankernel)` for more information.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  kernel_size = 10,
  batch_wise = FALSE,
  ref_qc_types = "SPL",
  recalc_trend_after = TRUE)

```

Again, we inspect the data before and after the correction. We observe that the green trendline
follows the samples (in grey). After the correction, the trend has been fully smoothed out,
resulting in a straight line.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

Now, the trends of the study samples appear to be fairly well corrected. However, in this example,
the differences compared to the previous QC-based smoothing are not very pronounced.

## Within-batch drift correction

In the examples before we app;ied drift correction across all batches. However, if batch-effects are
present that break the drifts, it is recommended to apply drift correction within each batch.

We explore a within-batch drift correction using a sample-based gaussian kernel smoothing from above
in this next example, by setting `batch_wise = TRUE`.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  kernel_size = 10,
  batch_wise = TRUE,
  ref_qc_types = "SPL",
  recalc_trend_after = TRUE)
```

Now, each batch has its own trendline corresponding the trend in each batch of each feature.
Contrary to the previous correction across batches, we now observe clear differences in the trends
still present after the correction. This is an effect of the correct applied independly to each
batch, to different sample sizes and present batch effects. Therefore, it is often necessary to
apply batch correction after bathc-wise drift correction

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

We therefore apply a subsequent batch correction using median-centering, resulting in an alignment
of the batches.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
mexp_drift <- midar::correct_batch_centering(
  mexp_drift, 
  ref_qc_types = "SPL", 
  variable = "conc",
  correct_scale = TRUE
)

plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

## Export corrected data

Next, we can either continue to work with the corrected data using `MiDAR` functions or export the
data.

```{r}
save_dataset_csv(
  mexp_drift, 
  path = "drift-batch-corrected-conc-data.csv", 
  variable = "conc", 
  filter_data = FALSE
  )
```
